Using generalized additive models to analyze biomedical non-linear longitudinal data

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.tex
References do not work

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{lineno}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{graphicx}

%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}

%add hyperlink to the box
\newcommand{\orcidaffila}[1]{%
  \href{https://orcid.org/0000-0002-6014-4538}{\usebox{\myorcidaffilbox}}}

\newcommand{\orcidaffilb}[1]{%
  \href{https://orcid.org/0000-0002-6135-8191}{\usebox{\myorcidaffilbox}}}

%command for the package lineno
\linenumbers

%authors
\author[1]{Ariel I. Mundo \orcidaffila{}}
\author[2]{John R. Tipton \orcidaffilb{}}
\author[1]{Timothy J. Muldoon \thanks{Corresponding author, tmuldoon@uark.edu}}
\affil[1]{\footnotesize Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
\affil[2]{\footnotesize Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}

%theme colors for the code chunks (originally from latex-solarized on GitHub)
%https://github.com/jez/latex-solarized
\usepackage{xcolor}
\definecolor{sbase03}{HTML}{002B36}
\definecolor{sbase02}{HTML}{073642}
\definecolor{sbase01}{HTML}{586E75}
\definecolor{sbase00}{HTML}{657B83}
\definecolor{sbase0}{HTML}{839496}
\definecolor{sbase1}{HTML}{93A1A1}
\definecolor{sbase2}{HTML}{EEE8D5}
\definecolor{sbase3}{HTML}{FDF6E3}
\definecolor{syellow}{HTML}{B58900}
\definecolor{sorange}{HTML}{CB4B16}
\definecolor{sred}{HTML}{DC322F}
\definecolor{smagenta}{HTML}{D33682}
\definecolor{sviolet}{HTML}{6C71C4}
\definecolor{sblue}{HTML}{268BD2}
\definecolor{scyan}{HTML}{2AA198}
\definecolor{sgreen}{HTML}{859900}
%command to set parameter(s) in package listings
\lstset{
    % How/what to match
    sensitive=true,
    % Border (above and below)
    frame=lines,
    % Extra margin on line (align with paragraph)
    xleftmargin=\parindent,
    % Put extra space under caption
    belowcaptionskip=1\baselineskip,
    % Colors
    backgroundcolor=\color{sbase3},
    basicstyle=\color{sbase00}\ttfamily,
    keywordstyle=\color{scyan},
    commentstyle=\color{sbase1},
    stringstyle=\color{sblue},
    numberstyle=\color{sviolet},
    identifierstyle=\color{sbase00},
    % Break long lines into multiple lines?
    breaklines=true,
    % Show a character for spaces?
    showstringspaces=false,
    tabsize=2
}

%\lstset{
%  breaklines=true,
%  stringstyle=\ttfamily,
%  backgroundcolor=\color{gray}
%}
\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\usepackage{float}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.

In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

@@

Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others. Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of the different situations where researchers have used longitudinal designs to study some physiological response. Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.

Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.

The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.

Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study. Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}. Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.

On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}. Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}. In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.

Due to a variety of causes, the number of observations during a study can vary between all subjects. For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible. It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points. In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated. When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}. This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}. Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}. And even more importantly, LMEMs also assume a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}, making them unsuitable to analyze non-linear data.

As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference. In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data. Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]} . There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]} and 3) can easily accommodate missing observations. Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.

The current advances in programming languages designed for statistical analysis (specifically \(\textsf{R}\)), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs. In particular, \(\textsf{R}\) {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, \(\textsf{R}\) has many tools that simplify data simulation, an emerging strategy used to test statistical models {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}. Data simulation methods allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.

This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of LMEMs and rm-ANOVA regarding linearity of response, constant correlation structures and missing observations are explained in detail. Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations. Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}. The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies. Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in \(\textsf{R}\), in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.

@@

\FloatBarrier

@@

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
\label{eq:linear-model}
\end{equation}

In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively. Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'' or ``saline'' vs.~``chemotherapy''). Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B). With this notation, the linear model then can be expressed as

>>>>

\documentclass[Royal,times,sagev]{sagej}

\usepackage{moreverb,url,natbib, multirow, tabularx}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

% Pandoc citation processing

\begin{document}

\title{Using generalized additive models to analyze biomedical
non-linear longitudinal data}

\runninghead{Mundo \emph{et al}.}

\author{Ariel I. Mundo\affilnum{1}, John R. Tipton\affilnum{2}, Timothy
J. Muldoon*\affilnum{1}}

\affiliation{\affilnum{1}{Department of Biomedical Engineering,
University of Arkansas, Fayetteville, AR, US}\\\affilnum{2}{Department
of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US}}

\corrauth{Corresponding author}

\email{\href{mailto:tmuldoon@uark.edu}{\nolinkurl{tmuldoon@uark.edu}}}

\begin{abstract}
In biomedical research, the outcome of longitudinal studies has been
traditionally analyzed using the \emph{repeated measures analysis of
variance} (rm-ANOVA) or more recently, \emph{linear mixed models}
(LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of
correlation and missing observations, both methodologies share an
assumption of linearity in the measured response, which results in
biased estimates and unreliable inference when they are used to analyze
data where the trends are non-linear, which is a common occurrence in
biomedical research. In contrast, \emph{generalized additive models}
(GAMs) relax the linearity assumption, and allow the data to determine
the fit of the model while permitting missing observations and different
correlation structures. Therefore, GAMs present an excellent choice to
analyze non-linear longitudinal data in the context of biomedical
research. This paper summarizes the limitations of rm-ANOVA and LMEMs
and uses simulated data to visually show how both methods produce biased
estimates when used on non-linear data. We also present the basic theory
of GAMs, and using trends of oxygen saturation in tumors reported in the
biomedical literature, we simulate example longitudinal data (2
treatment groups, 10 subjects per group, 5 repeated measures for each
group) to demonstrate how these models can be computationally
implemented. We show that GAMs are able to produce estimates that are
consistent with the trends of biomedical non-linear data even in the
case when missing observations exist (with 40\% of the simulated
observations missing), allowing reliable inference from the data. To
make this work reproducible, the code and data used in this paper are
available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.
\end{abstract}

\keywords{Cancer biology; Tumor response; generalized additive models;
simulation; R}

\maketitle

@@

Longitudinal studies are designed to repeatedly measure a variable of
interest in a group (or groups) of subjects, with the intention of
observing the evolution of effect across time rather than analyzing a
single time point (e.g., a cross-sectional study). Biomedical research
frequently uses longitudinal studies to analyze the evolution of a
``treatment'' effect across multiple time points; and in such studies
the subjects of analysis range from animals (mice, rats, rabbits), to
human patients, cells, or blood samples, among many others. Tumor
response \citep{roblyer2011, tank2020, pavlov2018, demidov2018},
antibody expression \citep{ritter2001, roth2017}, and cell metabolism
\citep{jones2018, skala2010} are examples of the different situations
where researchers have used longitudinal designs to study some
physiological response. Because the frequency of the measurements in a
longitudinal study is dependent on the biological phenomena of interest
and the experimental design of the study, the frequency of such
measurements can range from minute intervals to study a short-term
response such as anesthesia effects in animals\citep{greening2018}, to
weekly measurements to analyze a mid-term response like the evolution of
dermatitis symptoms in breast cancer patients \citep{sio2016}, to
monthly measurements to study a long-term response such as mouth opening
following radiotherapy (RT) in neck cancer patients \citep{kamstra2015}.

Traditionally, a ``frequentist'' or ``classical'' statistical paradigm
is used in biomedical research to derive inferences from a longitudinal
study. The frequentist paradigm regards probability as the limit of the
expected outcome when an experiment is repeated a large number of times
\citep{wagenmakers2008}, and such view is applied to the analysis of
longitudinal data by assuming a null hypothesis under a statistical
model that is often an \emph{analysis of variance over repeated
measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model
makes three key assumptions regarding longitudinal data: 1) linearity of
the response across time, 2) constant correlation across same-subject
measurements, and 3) observations from each subject are obtained at all
time points through the study (a condition also known as \emph{complete
observations}) \citep{gueorguieva2004, schober2018}.

The expected linear behavior of the response through time is a key
requisite in rm-ANOVA \citep{pinheiro2006}. This ``linearity
assumption'' in rm-ANOVA implies that the model is misspecified when the
data does not follow a linear trend, which results in unreliable
inference. In biomedical research, non-linear trends are the norm rather
than the exception in longitudinal studies. A particular example of this
non-linear behavior in longitudinal data arises in measurements of tumor
response to chemo and/or radiotherapy in preclinical and clinical
settings \citep{roblyer2011, skala2010, vishwanath2009}. These studies
have shown that the collected signal does not follow a linear trend over
time, and presents extreme variability at different time points, making
the fit of rm-ANOVA model inconsistent with the observed variation.
Therefore, when rm-ANOVA is used to draw inference of such data the
estimates are inevitably biased, because the model is only able to
accommodate linear trends that fail to adequately represent the
biological phenomenon of interest.

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to
perform repeated comparisons to estimate a \emph{p-value}, which in turn
is used as a measure of significance. Although it is possible that a
\emph{post hoc} analysis of rm-ANOVA is able to find ``significant''
\emph{p-values}( \emph{p}\textless0.05) from non-linear data, the
validity of such metric is dependent on how adequate the model fits the
data. In other words, \emph{p-values} are valid only if the model and
the data have good agreement; if that is not the case, a ``Type III''
error (known as ``model misspecification'') occurs\citep{dennis2019}.
For example, model misspecification will occur when a model that is only
able to explain linear responses (such as rm-ANOVA) is fitted to data
that follows a quadratic trend, thereby causing the resulting
\emph{p-values} and parameter estimates to be invalid \citep{wang2019}.

Additionally, the \emph{p-value} itself is highly variable, and multiple
comparisons can inflate the false positivity rate (Type I error or
\(\alpha\)) \citep{liu2010, halsey2015}, consequently biasing the
conclusions of the study. Corrections exist to address the Type I error
issue of multiple comparisons (such as Bonferroni \citep{abdi2010}), but
they in turn reduce statistical power (1-\(\beta\))\citep{nakagawa2004},
and lead to increased Type II error (failing to reject the null
hypothesis when the null hypothesis is false)
\citep{gelman2012, albers2019}. Therefore, the tradeoff of \emph{post
hoc} comparisons in rm-ANOVA between Type I, II and III errors might be
difficult to resolve in a biomedical longitudinal study where a delicate
balance exists between statistical power and sample size.

On the other hand, the assumption of constant correlation in rm-ANOVA
(often known as the \emph{compound symmetry assumption}) is typically
unreasonable because correlation between the measured responses often
diminishes as the time interval between the observation increases
\citep{ugrinowitsch2004}. Corrections can be made in rm-ANOVA in the
absence of compound symmetry \citep{huynh1976, greenhouse1959}, but the
effectiveness of the correction is limited by the size of the sample,
the number of measurements\citep{haverkamp2017}, and group sizes
\citep{keselman2001}. In the case of biomedical research, where living
subjects are frequently used, sample sizes are often not ``large'' due
to ethical and budgetary reasons \citep{charan2013} which might cause
the corrections for lack of compound symmetry to be ineffective.

Due to a variety of causes, the number of observations during a study
can vary between all subjects. For example, in a clinical trial patients
may voluntarily withdraw, whereas attrition due to injury or weight loss
in preclinical animal studies is possible. It is even plausible that
unexpected complications with equipment or supplies arise that prevent
the researcher from collecting measurements at certain time points. In
each of these missing data scenarios, the \emph{complete observations}
assumption of classical rm-ANOVA is violated. When incomplete
observations occur, a rm-ANOVA model is fit by excluding all subjects
with missing observations from the analysis \citep{gueorguieva2004}.
This elimination of partially missing data from the analysis can result
in increased costs if the desired statistical power is not met with the
remaining observations, because it would be necessary to enroll more
subjects. At the same time, if the excluded observations contain
insightful information that is not used, their elimination from the
analysis may limit the demonstration of significant differences between
groups.

During the last decade, the biomedical community has started to
recognize the limitations of rm-ANOVA in the analysis of longitudinal
data. The recognition on the shortcomings of rm-ANOVA is exemplified by
the use of linear mixed effects models (LMEMs) by certain groups to
analyze longitudinal tumor response data
\citep{skala2010, vishwanath2009}. Briefly, LMEMs incorporate
\emph{fixed effects}, which correspond to the levels of experimental
factors in the study (e.g., the different drug regimens in a clinical
trial), and \emph{random effects}, which account for random variation
within the population (e.g., the individual-level differences not due to
treatment such as weight or age). When compared to the traditional
rm-ANOVA, LMEMs are more flexible as they can accommodate missing
observations for multiple subjects and allow different modeling
strategies for the variability within each measure in every subject
\citep{pinheiro2006}. However, LMEMs impose restrictions in the
distribution of the errors of the random effects, which need to be
normally distributed and independent \citep{gueorguieva2004, barr2013}.
And even more importantly, LMEMs also assume a linear relationship
between the response and time \citep{pinheiro2006}, making them
unsuitable to analyze non-linear data.

As the rm-ANOVA and the more flexible LMEM approaches make overly
restrictive assumptions regarding the linearity of the response, there
is a need for biomedical researchers to explore the use of additional
statistical tools that allow the data (and not an assumption in trend)
to determine the trend of the fitted model, to enable appropriate
inference. In this regard, generalized additive models (GAMs) present an
alternative approach to analyze longitudinal data. Although not
frequently used by the biomedical community, these semi-parametric
models are customarily used in other fields to analyze longitudinal
data. Examples of the use of GAMs include the analysis of temporal
variations in geochemical and palaeoecological data
\citep{rose2012, pedersen2019, simpson2018}, health-environment
interactions \citep{yang2012} and the dynamics of government in
political science \citep{beck1998} . There are several advantages of
GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible
class of smooth responses that enable the data to dictate the trend in
the fit of the model, 2) they can model non-constant correlation between
repeated measurements \citep{wood2017} and 3) can easily accommodate
missing observations. Therefore, GAMs can provide a more flexible
statistical approach to analyze non-linear biomedical longitudinal data
than LMEMs and rm-ANOVA.

The current advances in programming languages designed for statistical
analysis (specifically \(\textsf{R}\)), have eased the computational
implementation of traditional models such as rm-ANOVA and more complex
approaches such as LMEMs and GAMs. In particular, \(\textsf{R}\)
\citep{r} has an extensive collection of documentation and functions to
fit GAMs in the package \emph{mgcv} \citep{wood2016, wood2017} that not
only speed up the initial stages of the analysis but also enable the use
of advanced modeling structures (e.g.~hierarchical models, confidence
interval comparisons) without requiring advanced programming skills from
the user. At the same time, \(\textsf{R}\) has many tools that simplify
data simulation, an emerging strategy used to test statistical models
\citep{haverkamp2017}. Data simulation methods allow the researcher to
create and explore different alternatives for analysis without
collecting information in the field, reducing the time window between
experiment design and its implementation, and simulation can be also
used for power calculations and study design questions.

This work provides biomedical researchers with a clear understanding of
the theory and the practice of using GAMs to analyze longitudinal data
using by focusing on four areas. First, the limitations of LMEMs and
rm-ANOVA regarding linearity of response, constant correlation
structures and missing observations are explained in detail. Second, the
key theoretical elements of GAMs are presented using clear and simple
mathematical notation while explaining the context and interpretation of
the equations. Third, we illustrate the type of non-linear longitudinal
data that often occurs in biomedical research using simulated data that
reproduces patterns in previously reported studies
\citep{vishwanath2009}. The simulated data experiments highlight the
differences in inference between rm-ANOVA, LMEMs and GAMs on data
similar to what is commonly observed in biomedical studies. Finally,
reproducibility is emphasized by providing the code to generate the
simulated data and the implementation of different models in
\(\textsf{R}\), in conjunction with a step-by-step guide demonstrating
how to fit models of increasing complexity.

@@

@@

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times time_{t} +\beta_2 \times treatment_{j} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt}\\ 
(\#eq:linear-model)
\end{equation}

In this model \(y_{ijt}\) is the response for subject \(i\), in
treatment group \(j\) at time \(t\), which can be decomposed in a mean
value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment
(\(treatment_j\)) and their interaction \(time_t*treatment_j\) which
have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\),
respectively. Independent errors \(\varepsilon_{ijt}\) represent random
variation not explained by the \emph{fixed} effects, and are assumed to
be \(\sim N(0,\sigma^2)\) (independently and identically normally
distributed with mean zero and variance \(\sigma^2\)). In a biomedical
research context, suppose two treatments groups are used in a study
(e.g., ``placebo'' vs.~``novel drug'' or ``saline''
vs.~``chemotherapy''). Then, the group terms in Equation
@ref(eq:linear-model) can be written as below with \(treatment_j=0\)
representing the first treatment group (Group A) and \(treatment_j=1\)
representing the second treatment group (Group B). With this notation,
the linear model then can be expressed as
